Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-05-18 paper/paper.qmd
Report preprocess effects on OTU numbers
other minor tweaks

@@

This study was approved by the University of Michigan Institutional Review
Board. All patient samples were collected by the University of Michigan Health
System from January 2016 through December 2017.
Stool samples that had unformed stool consistency were tested for *C. difficile* by
the clinical microbiology lab with a two-step algorithm that included detection
of *C. difficile* glutamate dehydrogenase and toxins A and B by enzyme
immunoassay with reflex to PCR for the *tcdB* gene when results were discordant.
1,517 stool samples were collected from patients diagnosed with a CDI.
Leftover stool samples that were sent to the clinical microbiology lab were
collected and split into different aliquots.
For 16S sequencing, the aliquot of stool was resuspended in
DNA genotek stabilization buffer and then stored in the -80&deg;C freezer.
Only the first CDI sample per patient was used for subsequent ML analyses such
that no patient is represented more than once, resulting in a dataset of 1,191
samples.

@@

All sequences were processed with mothur (v1.46) using the MiSeq SOP protocol
[@schloss_introducing_2009; @kozich_development_2013].
Paired sequencing reads were combined and
aligned with the SILVA (v132) reference database [@quast_silva_2013] and taxonomy was
assigned with a modified version of the Ribosomal Database Project reference
sequences (v16) [@cole_ribosomal_2014].
Sequences were clustered into _de novo_ OTUs with the OptiClust algorithm in
mothur [@westcott_opticlust_2017].
Samples were rarefied to 5,000 sequences per sample, repeated 1,000
times for alpha and beta diversity analysis.
<!-- TODO supplementary figure with alpha and beta diversity & significance -->

@@

Random forest models were used to examine whether OTU data collected on the day
of diagnosis could classify CDI cases as severe according to four different
definitions of severity.
We used the mikropml R package v1.5.0 [@topcuoglu_mikropml_2021]
for all steps of the machine learning analysis.
We randomly split the data into an 80% training and 20% test set and repeated
this 100 times, followed by training models with 5-fold cross-validation.
Model performance was calculated on the test set using the area under the
receiver-operator characteristic curve (AUROC) and the area under the balanced
precision-recall curve (AUBPRC).
Permutation feature importance was then performed to determine which OTUs
contributed most to model performance.
We reported OTUs with a significant permutation test in at least 80 of the 100
models.

@@

The workflow was defined with Snakemake [@koster_snakemake_2012] using a custom
version of the mikropml Snakemake workflow [@sovacool_mikropml_2023].
Dependencies were managed with conda environments.
Scripts were written in R [@r_core_team_r_2020],
Python [@van_rossum_python_2009],
and GNU bash.
Additional software and packages used in the creation of this manuscript include
cowplot [@wilke_cowplot_2020],
ggtext [@wilke_ggtext_2020],
ggsankey [@sjoberg_ggsankey_2022],
schtools [@sovacool_schtools_2022],
the tidyverse metapackage [@wickham_welcome_2019],
Quarto, and
vegan [@oksanen_vegan_2023].

@@

::: {#fig-flowchart fig-cap="**CDI severity definitions.**
**A)** Decision flow chart to define CDI cases as severe according to the Infectious
Diseases Society of America (IDSA) based on lab values,
the occurence of complications due to any cause (All-cause),
and the occurence of disease-related complications confirmed as attributable to
CDI with chart review (Attrib).
**B)** The proportion of severe CDI cases labelled according to each definition.
An additional 'Pragmatic' severity definition uses the Attributable definition
when possible, and falls back to the All-cause definition when chart review is
not available.
<!-- TODO table (supplementary?) showing counts & frequency of positives-->
"}
![](figures/flowchart_sankey.png)
:::

@@

::: {#fig-features fig-cap="**Feature importance.**
Feature importance via permutation test.
For each OTU, the order of samples was randomized in the test set 100 times and
the performance was re-calculated to estimate the permutation performance.
An OTU was considered important if the performance decreased when the OTU was
permuted in at least 80% of the models.
OTUs with a greater difference in AUROC (actual performance minus permutation
performance) are more important.
Left: models were trained on the full dataset, with different numbers of samples
available for each severity definition.
Right: models were trained on the intersection of samples with all labels
available for each definition. Note that Attributable and Pragmatic severity are
exactly the same for the intersection set.
"}
![](figures/feature-importance.png)
:::

>>>>

```{r load_data}
library(here)
library(tidyverse)

load(here("results", "stats.RData"))
```

@@

This study was approved by the University of Michigan Institutional Review
Board. All patient samples were collected by the University of Michigan Health
System from January 2016 through December 2017.
Stool samples that had unformed stool consistency were tested for *C. difficile* by
the clinical microbiology lab with a two-step algorithm that included detection
of *C. difficile* glutamate dehydrogenase and toxins A and B by enzyme
immunoassay with reflex to PCR for the *tcdB* gene when results were discordant.
1,517 stool samples were collected from patients diagnosed with a CDI.
Leftover stool samples that were sent to the clinical microbiology lab were
collected and split into different aliquots.
For 16S sequencing, the aliquot of stool was resuspended in
DNA genotek stabilization buffer and then stored in the -80&deg;C freezer.
Only the first CDI sample per patient was used for subsequent ML analyses such
that no patient is represented more than once, resulting in a dataset of
`r n_cases_first` samples.

@@

All sequences were processed with mothur (v1.46) using the MiSeq SOP protocol
[@schloss_introducing_2009; @kozich_development_2013].
Paired sequencing reads were combined and
aligned with the SILVA (v132) reference database [@quast_silva_2013] and taxonomy was
assigned with a modified version of the Ribosomal Database Project reference
sequences (v16) [@cole_ribosomal_2014].
Sequences were clustered into _de novo_ OTUs with the OptiClust algorithm in
mothur [@westcott_opticlust_2017], resulting in 
Samples were rarefied to 5,000 sequences per sample, repeated 1,000
times for alpha and beta diversity analysis.
<!-- TODO supplementary figure with alpha and beta diversity & significance -->

@@

Random forest models were used to examine whether OTU data collected on the day
of diagnosis could classify CDI cases as severe according to four different
definitions of severity.
We used the mikropml R package v1.5.0 [@topcuoglu_mikropml_2021] implemented in
a custom version of the mikropml Snakemake workflow [@sovacool_mikropml_2023]
for all steps of the machine learning analysis.
We have a full dataset which uses all samples available for each severity
definition, and an intersection dataset which consists of only the samples that
have all four definitions labelled.
The intersection dataset is the most fair for comparing model performance across
definitions, while the full dataset allows us to use as much data as possible
for model training and evaluation.
Datasets were preprocessed with the default options in mikropml to remove
features with near-zero variance and scale continuous features from -1 to 1. <!-- No features had missing values and no features were perfectly correlated -->
During preprocessing, `r preproc_ranges$min_removed` to `r preproc_ranges$max_removed`
features were removed due to having near-zero variance, resulting in datasets
ranging from `r preproc_ranges$min_feats` to `r preproc_ranges$max_feats`
depending on the severity definition.
We randomly split the data into an 80% training and 20% test set and repeated
this 100 times, followed by training models with 5-fold cross-validation.
Model performance was calculated on the test set using the area under the
receiver-operator characteristic curve (AUROC) and the area under the balanced
precision-recall curve (AUBPRC).
Permutation feature importance was then performed to determine which OTUs
contributed most to model performance.
We reported OTUs with a significant permutation test in at least 80 of the 100
models.

@@

The workflow was defined with Snakemake [@koster_snakemake_2012] and
dependencies were managed with conda environments.
Scripts were written in R [@r_core_team_r_2020],
Python [@van_rossum_python_2009],
and GNU bash.
Additional software and packages used in the creation of this manuscript include
cowplot [@wilke_cowplot_2020],
ggtext [@wilke_ggtext_2020],
ggsankey [@sjoberg_ggsankey_2022],
schtools [@sovacool_schtools_2022],
the tidyverse metapackage [@wickham_welcome_2019],
Quarto, and
vegan [@oksanen_vegan_2023].

@@

::: {#fig-flowchart fig-cap="**CDI severity definitions.**
**A)** Decision flow chart to define CDI cases as severe according to the Infectious
Diseases Society of America (IDSA) based on lab values,
the occurence of an adverse outcome due to any cause (All-cause),
and the occurence of disease-related complications confirmed as attributable to
CDI with chart review (Attrib).
**B)** The proportion of severe CDI cases labelled according to each definition.
An additional 'Pragmatic' severity definition uses the Attributable definition
when possible, and falls back to the All-cause definition when chart review is
not available.
<!-- TODO table (supplementary?) showing counts & frequency of positives-->
"}
![](figures/flowchart_sankey.png)
:::

@@

::: {#fig-features fig-cap="**Feature importance.**
Feature importance via permutation test.
For each OTU, the order of samples was randomized in the test set 100 times and
the performance was re-calculated to estimate the permutation performance.
An OTU was considered important if the performance decreased when the OTU was
permuted in at least 75% of the models.
OTUs with a greater difference in AUROC (actual performance minus permutation
performance) are more important.
Left: models were trained on the full dataset, with different numbers of samples
available for each severity definition.
Right: models were trained on the intersection of samples with all labels
available for each definition. Note that Attributable and Pragmatic severity are
exactly the same for the intersection set.
OTU 120 (*Pseudomonas*) is not shown for the full data set with IDSA severity 
on the full dataset because it was removed during pre-processing due to having
near-zero variance.
"}
![](figures/feature-importance.png)
:::
